---
title: "Juvenile Snowshoe Hares in Bonanza Creek"
author: "Van Gonzales"
date: "11/3/2021"
output: 
html_document:
  toc: yes
  theme: paper
  number_sections: yes 
---
```{r setup, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.align = 'center', out.extra = 'angle = 90')
library(tidyverse) # Loading all packages
library(here)
library(janitor)
library(kableExtra)
library(broom)
library(lubridate)
library(extrafont)
library(effsize)
library(equatiomatic)
library(plotly)
library(effsize)
```

## Introduction
This report provides an exploratory analysis of juvenile snowshoe hares in the Bonanza Creek forest in Tanana Valley Alaska. The data was observed and recorded from 1999 - 2012. The hares were observed in three different locations: Riparian Forest (bonrip), Mature Forest (bonmat), Lowlands Black Spruce Forest (bonbs). Bonrip is a flood zone, Bonmat is a mature black spruce forest, and Bonbs was the site used for live trapping the snowshoe hares for monitoring. The following report includes an analysis of the number of juvenile hare trappings conducted each year of the study, an analysis of the relationship between site and gender on average hare weight, and linear regression analysis of the relationship between hare weight and the hind foot length. 

## Data and methods
Dr. Knut Kielland (Kielland et. al. 2017.) was the principal investigator for this project. This dataset includes information on the trappings of both adult and juvenile hares at all 3 sites including date, trap number, and ear tag identifier. In addition, the dataset contains information on each hare trapped, including sex, weight, hind foot length, and age. In this report, a two-sample unpaired t-test with an $\alpha$ = 0.05 was used to compare the average weights between male and female juvenile hares. Differences between groups are described by Cohen's *d* effect size. A simple linear regression was used to model the relationship between hare weight and hind foot length. All analyses are on R Version 4.1.1 using RStudio Version 1.4.1717.
<br>
```{r}
bonanza_hares <- read_csv(here("data", "bonanza_hares.csv")) #Reading in the Data
```
## Juvenile Hare Investigation
<br>

#### A. Trap Counts Per Year
<br>
```{r}
trap_counts <- bonanza_hares %>% 
  filter(age == "j") %>%  #Filtering for just juveniles
  mutate(date = mdy(date)) %>% # Changing the date
  mutate(year = year(date)) %>% 
  group_by(year) %>% 
  summarize(trap = n()) #Summarizing the trap count
```


```{r}
trap_plot <- ggplot(data = trap_counts, aes(x = year, y = trap)) +
  geom_col(fill = "blue2") +
  scale_x_continuous(breaks = seq(from = 1999, to = 2012, by = 1)) +
  labs(x = "Year", y = "# of Trappings", 
       title = "Trapping Count of Juvenile Hares by Year") +
   theme(axis.text.x = element_text(family = "Times New Roman", #family changes font
                size = 7),
                axis.text.y = element_text(family = "Times New Roman",
                                          size = 8),
         plot.title = element_text(hjust = 0.5, family = "Times New Roman"), #hjust centers the title
         axis.title.x = element_text(family = "Times New Roman"),
         axis.title.y = element_text(family = "Times New Roman"))

ggplotly(trap_plot)  
```
**Figure 1.** Histogram of juvenile hare trappings in Bonanza Creek from 1998-2012, from Kielland et al. (2017). Over the duration of the study, the trappings per year remained inconsistent.  

```{r}
trap_summary <- trap_counts %>% 
  summarize(mean_trap = mean(trap), # creating columns for median, mean, max, and min
            median_trap = median(trap),
            max_trap = max(trap),
            min_trap = min(trap))
```
*Trappings Analysis*
<br>
The minimum number of trappings each year, `r trap_summary$min_trap`, occurred in 2002, and the maximum number of trappings each year, `r trap_summary$max_trap`, occurred in 2009. The average amount of trappings performed each year over the entire project was `r trap_summary$mean_trap` and the median was `r trap_summary$median_trap`. There does not seem to be a consistent or trending number of trappings from year to year. There are several factors that could be affecting the trapping rates such as the number of days and the number of traps each year in the study. This could be due to researcher effort, difficulty trapping younger hares, or even seasonal variability in juvenile hare presence in those locations. In order to standardize the juvenile hare population moving forward, researchers should conduct the study at the same time each year, and place the same number of traps from year to year.
<br>

#### B. Visualize Hare Weights By Sex & Site

```{r}
weight_sex_site <- bonanza_hares %>% 
  filter(age == "j") %>% 
  group_by(sex, grid) %>% 
  summarize(mean_weight = round(mean(weight, na.rm = TRUE), 2),
            median_weight = round(median(weight, na.rm = TRUE), 2),
            sd_weight = round(sd(weight, na.rm = TRUE), 2)) %>% 
  mutate(grid = case_when(
    grid == "bonrip" ~ "Bonanza Riparian", # changing the facet names
    grid == "bonbs" ~ "Bonanza Lowland",
    grid == "bonmat" ~ "Bonanza Mature"))
```

```{r}
weight_sex_site_plot<- ggplot(data = weight_sex_site, 
       aes(x = sex, y = mean_weight, fill = sex)) +
  geom_bar(stat = "identity") + # stat identity when choosing x and y variables
  facet_wrap(~grid) + # wrapping by site
  theme(strip.text.x = element_text(family = "Times New Roman"))+
  scale_fill_manual(values = c("palevioletred1", "slateblue3")) + # manual selection of colors
  labs(x = "Sex", y = "Average Weight (g)",
       title = "Average Weight of Male and Female Hares by Site") +
  theme(axis.text.x = element_text(family = "Times New Roman", # cahnging font
                size = 9),
                axis.text.y = element_text(family = "Times New Roman",
                                          size = 8),
         plot.title = element_text(hjust = 0.5, family = "Times New Roman"),
         axis.title.x = element_text(family = "Times New Roman"),
         axis.title.y = element_text(family = "Times New Roman"),
        ) +
  geom_errorbar(aes(ymin = mean_weight - sd_weight, ymax = mean_weight + sd_weight),
                width=.2, position=position_dodge(.9)) # adding sd error bars

ggplotly(weight_sex_site_plot) #making the plot interactive
```
**Figure 2.** Bar plot of male (purple), female (pink), and the unknown sex (gray) juvenile hare weights by site in Bonanza Creek from 1998-2012, from Kielland et al. (2017). The error bars are representative of the standard deviation of the sample weights. The unknown genders are represented by NA.
<br>

*Weight by Sex & Site Analysis*
<br>
As **Figure 2** shows, the female juvenile hares have a lower average weight compared to males at every location. At the Bonanza Lowland location, the females average weight is `r weight_sex_site$mean_weight[1]` $\pm$  `r weight_sex_site$sd_weight[1]` grams compared to the male average weight of  `r weight_sex_site$mean_weight[4]` $\pm$  `r weight_sex_site$sd_weight[4]` grams. The NA bar column does not show up in this site because there isn't enough data to calculate a mean weight. At the Bonanza Mature site,  the female average weight is `r weight_sex_site$mean_weight[2]` $\pm$  `r weight_sex_site$sd_weight[2]` grams compared to the male average weight of `r weight_sex_site$mean_weight[5]` $\pm$  `r weight_sex_site$sd_weight[5]` grams and the unknown sex hare average weight of `r weight_sex_site$mean_weight[8]` $\pm$  `r weight_sex_site$sd_weight[8]` grams. Finally, at the Bonanza Riparian site, the female average weight is `r weight_sex_site$mean_weight[3]` $\pm$  `r weight_sex_site$sd_weight[3]` grams compared to the male average weight of `r weight_sex_site$mean_weight[6]` $\pm$  `r weight_sex_site$sd_weight[6]`grams and the unknown sex hare average weight of `r weight_sex_site$mean_weight[9]` $\pm$  `r weight_sex_site$sd_weight[9]` grams. 

```{r, include = FALSE} 
juveniles_sex_weight <- bonanza_hares %>% 
  filter(age == "j") %>% 
  group_by(sex) %>% 
  summarize(weight) #creating df with just weight and sex

ggplot(data = juveniles_sex_weight, 
       aes(x = weight)) +
  geom_histogram() +
  facet_wrap(~sex)
```
```{r, include = FALSE}
ggplot(data = juveniles_sex_weight, aes(sample = weight)) +
  stat_qq() + #creating qq plot
  geom_qq_line() +
  facet_wrap(~sex)
```

```{r}
f_hares <- juveniles_sex_weight %>% 
  filter(sex == "f") %>% 
  drop_na(weight) %>% #removing na from weight
  pull(weight)

m_hares <- juveniles_sex_weight %>% 
  filter(sex == "m") %>% 
  drop_na(weight) %>% 
  pull(weight) # pulling just weight from juveniles_sex_weight df

```

```{r}
hares_t <- t.test(f_hares, m_hares) #performing ttest
```

```{r}
hare_cohen <- cohen.d(f_hares, m_hares) #performing cohen d test
```


```{r}
weight_sex_count <- bonanza_hares %>% 
  filter(age == "j") %>% 
  group_by(sex) %>% 
  summarize(mean_weight = round(mean(weight, na.rm = TRUE), 1),
            sd_weight = round(sd(weight, na.rm = TRUE), 1),
            sample_size = n(),) %>% 
  na.omit() #removing all nas

mean_diff = weight_sex_count$mean_weight[2] - weight_sex_count$mean_weight[1] #calc mean difference
```

#### C. Juvenile Weight Comparison Between Male and Female Hares
```{r}
weight_sex_count %>% 
  kable(col.names = c("Sex", #creating column names
                      "Mean Weight (g)",
                      "Standard Deviation(g)",
                      "Sample Size (n)"),
                      caption = "Table 1. Comparing juvenile hare average weights between males and females in Bonanza Creek from 1998-2012, from Kielland et al. (2017). The male mean weight is larger than the female mean weight of all juveniles across all sites.") %>% #including caption in table
  kable_classic_2(full_width = T, html_font = "Cambria") #table styling
```
<br>

*Weight by Sex Analysis (NAs Excluded)*
<br>
The actual difference in these means is `r round(mean_diff,1)` grams. The mean weight of female hares across all three sites is `r weight_sex_count$mean_weight[1]` $\pm$ `r weight_sex_count$sd_weight[1]` grams and the male hares average weight across all sites is `r weight_sex_count$mean_weight[2]` $\pm$ `r weight_sex_count$sd_weight[2]` grams. A standard unpaired two sample t-test was conducted to analyze if the calculated difference is statistically significant. The test resulted in a p-value of `r round(hares_t$p.value, 3)` (effect size, `r abs(round(hare_cohen$estimate,4))`), and means there is a `r round(hares_t$p.value * 100, 3)`% chance that these samples would have a difference in means at least as different as `r round(mean_diff, 1)` grams, assuming that the two samples are drawn from populations with the same mean. The null hypothesis is that the difference in the means between males and females is equal to 0, while the alternative hypothesis is that the difference in means is not equal to 0. Therefore with this p-value, the null hypothesis is rejected because it falls below the assigned significance level of ($\alpha$ = 0.05).
<br>

#### D. Relationship Between Juvenile Weight & Hind Foot Length
```{r}
juveniles <- bonanza_hares %>% 
  filter(age == "j") #creating df with just J's should've done this at beginning
```


```{r}
hindft_plot <- ggplot(data = juveniles, 
                      aes(y = weight, x = hindft)) +
  geom_point() +
  geom_smooth(method = "lm") + #adding linear regression line
  labs(y = "Weight (g)",
       x = "Hind Foot Length (mm)",
       title = "Juvenile Hares Hind Foot vs Weight") + 
    theme(axis.text.x = element_text(family = "Times New Roman", 
                size = 7),
                axis.text.y = element_text(family = "Times New Roman",
                                          size = 8),
         plot.title = element_text(hjust = 0.5, family = "Times New Roman"),
         axis.title.x = element_text(family = "Times New Roman"),
         axis.title.y = element_text(family = "Times New Roman"))
  
  ggplotly(hindft_plot) # making plot interactive
```
**Figure 3.** Linear regression model scatterplot showing length of juvenile weight (g) as a function of hind foot length (mm). The dots are representative of individual juvenile hare observations. The blue line plotted across the dots represents a simple linear regression. Data: (Kielland et al. 2017).


```{r, include = FALSE}
hare_pearson <- cor.test(juveniles$weight, juveniles$hindft) #pearson r test

hare_pearson_tidy <- tidy(hare_pearson) #cleaning up the t4est

hares_lm <- lm(weight ~hindft, data = juveniles) #linear regression model

hares_lm_glance <- glance(hares_lm) #lm table; call it in console

hares_lm_tidy <- tidy(hares_lm) #cleaning up lm

plot(hares_lm) #plotting lm; can also do in console

extract_eq(model = hares_lm, use_coefs = TRUE) #looking at coefficients

hares_lm_glance #calling lm table; can also do in console
```
*Linear Regression of Hind Foot & Weight Analysis*
<br>
The relationship between juvenile hare weight and hind foot length is described by the linear regression model displayed by **Figure 3**. The linear regression model suggests that hind foot length affects weight (p < 0.001, R^2^ = `r round(hares_lm_glance$r.squared,2)`) with the average slope of $\beta$ = `r round(hares_lm_tidy$estimate[2], 2)` g/mm. According to the Pearson's *r* test, the correlation between hind foot and weight is significant and positive (Pearson's *r* = `r round(hare_pearson_tidy$estimate,2)`, p < 0.001). Although these p-values suggest significance, it is concerning that the assumption of homoscedasticity in the linear regression is not satisfied. Therefore, performing a linear regression is not an appropriate method of determining the relationship between hind foot length and weight.

## Summary
- The trappings over the duration of the of the study were inconsistent year to year. 
- There was a large variation in the average weight of juvenile hares by sex and site, however overall male juvenile hares had larger weights compared to females and the unidentified sexes. The p-value for this comparison of male and female weight was found to be statistically significant.
- There is not a linear relationship between hare hind foot length and its weight.
- There is a statistically significant difference in means between male and female juvenile snowshoe hare weights and the effect size is small. 
- The total juvenile hare trap counts by year steadily declined from the beginning of sampling (1999) to the end (2012).


## Citations
Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017. Snowshoe hare physical data in Bonanza Creek Experimental Forest: 1999-Present ver 22. Environmental Data Initiative. https://doi.org/10.6073/pasta/03dce4856d79b91557d8e6ce2cbcdc14













