---
title: "Mini Report- Snowshoe Hares"
author: "Van Gonzales"
date: "11/3/2021"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE) #Change these two to false in the end (message and warning)
library(tidyverse)
library(here)
library(janitor)
library(kableExtra)
library(broom)
library(lubridate)
```

### 1. Introduction

This report provides an exploratory overview of the counts and sizes of juvenile snowshoe hares recorded from 1998-2012. (gentoo, chinstrap and Adélie) observed on islands within the Palmer Archipelago from 2007 - 2009 (Gorman et al. 2014). Specifically, flipper lengths are compared across the three species, and the relationship(s) between flipper length and body mass explored. NOT DONE YET

### 2. Data and methods

Penguin structural size measurements were collected and made available by Dr. Kristen Gorman and colleagues at the Palmer Station Long Term Ecological Research (LTER) network site in Palmer Archipelago, Antarctica (see map from Gorman et al. (2014) in Figure 1). The data contain observations for 344 unique nesting penguins, collected over three seasons (2007, 2008, 2009). Following exploratory data visualization, flipper lengths between male and female penguins are by two-sample t-tests using a significance level ($\alpha$) of 0.05 throughout. Differences between groups are described by Cohen's *d* effect size. The relationship between flipper length and body mass is explored by simple linear regression. All analyses are in R version 4.0.2 using RStudio version 1.3.1056.

```{r}
bonanza_hares <- read_csv(here("data", "bonanza_hares.csv"))
```

**Figure 1.** Histogram of juvenile hare trappings in Bonanza Creek from 1998-2012, from Kielland et al. (2017). 

### 3. Preliminary results

#### A. Junvenile hare trap counts during each year
```{r}
trap_counts <- bonanza_hares %>% 
  filter(age == "j") %>% 
  mutate(date = mdy(date)) %>%
  mutate(year = year(date)) %>% 
  group_by(year) %>% 
  summarize(trap = n())
```


```{r}
ggplot(data = trap_counts, aes(x = year, y = trap)) +
  geom_col(fill = "blue2") +
  scale_x_continuous(breaks = seq(from = 1999, to = 2012, by = 1)) +
  labs(x = "Year", y = "# of Trappings", 
       title = "Trapping Count of Juveniles by Year")
  
```

Then, in 3 - 4 sentences below your figure, describe major takeaways from your graph about total snowshoe hare trappings at the site. You might consider:
What were the minimum and maximum counts of juvenile hare trappings?
Mean or median annual number of juvenile hares trapped?
General/notable trends?
These are absolute counts, but this count is also impacted by effort (e.g. the number of days and number of traps in the study each year). What would you suggest be done moving forward (you are not expected to actually do this) to standardize the juvenile hare population in future analyses?





